Seismic Time Series Analysis\\Statistics for Spatio-Temporal Data\\\textit{-- project draft --}

### 2020-06-06 draft.tex
fix draft as Diletta and Fabio suggested

\subsubsection{Motivation and objective}
Thanks to the recent coronavirus outbreak worldwide, we currently have data about many different phoenomena from periods with standard human activity, and from a period with reduced human activity.

@@

It is important to notice that the \textit{day}ly data in each file is considered in UTC time, an optimal choice to publish data in a machine-readable way, even if this could cause some issues while interpreting the results.

@@

\subsubsection{Data aggregation and initial transformation}
Since we are not interested in second-precise analysis, and it would be computationally expensive to deal with the whole dataset, we are going to aggregate data by hour. The aggregation will get us around 14 thousand points, that can still result in interpretable models.

@@

In particular, after a visual analysis after decomposition, we can suppose the 90 hours gap from the Uccle station is related to some changes to the seismic equipment, that resulted in slightly more precise measurements.

@@

Since our main focus is on the seasonalities themselves, in order to prevent assumptions based on the objective, so that searching for correlation with human activity becomes imposing human-based seasonality periods, we will use an analytical method to extract seasonalities that the data itself suggests to us. Still, we recognize that the already performed transformations impose an human-based time alteration to the data itself (Section \ref{sec:cleanup-transform}).

@@

On this note we must specify that, when a significant periodicity is found, it influences smaller periodicities given by $f*\{2,3,\dots\}$ the significant frequency. Also, if any periodicity is found that is longer than the original dataset, it must be considered as a \textit{coincidence}, rather than a surprising serendipity.

@@

In \figurename~\ref{fig:decomposed} we see the decomposition results. While data from Uccle station have a clear significativity of both trend and seasonalities, data from Membach could seem not significant. Still, if we notice that the main timeseries lies between values of 0 and 100, we can find out that around half of the value will be composed by the trend, a 30\% by the joined seasonalities component, and the remaining value by the residuals. We will need to better handle the spikes from around half our dataset.

@@

@@

Since our dataset shows a relevant trend component, resulting in non-constant mean, we consider the usage of the \textit{diff} operator to smooth the mean.
After applying the \textit{diff} operator with lag 168, in conjunction with the simple \textit{diff} or not, we come up with two possible models by checking the ACF and PACF plots:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item from using both \textit{diff}s, the ACF and PACF on the long run hint to $\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$
	\item from using just the lagged \textit{diff}, the PACF on the first lags and the ACF/PACF combination on the long run hint to $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$
\end{itemize}
We also employ the \textit{auto.arima} stepwise methodology, locking the \textit{diff} values to 0 or 1, to get some hints on the possible models.

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=.7\linewidth]{project.uccs_files/figure-latex/arima-26} %30
	\caption{ACF and PACF plots for $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
	\label{fig:arima-26}
\end{figure}

The ACF and PACF of the residuals for all three models are really similar (\figurename~\ref{fig:arima-26}), even if the models themselves have great differences in meaning and in number of parameters: they highlight a clear seasonal ARMA process with periodicity 24. This shows how our previous conclusion that the 168 hours seasonality could include the 24 hours one is flawed. Unfortunately, the parameters estimator we are using does not permit to estimate multiple seasonalities\footnote{\textit{Arima} method from the \textit{forecast} R package, only specifying the seasonal and non-seasonal orders}.

Probably related to the 24 hours periodicity that remains on the residuals, together with the non-Normality of the original data, all Q-Q plots show clear divergence on both sides of the distribution. The Box-Pierce test statistics accept the independence hypothesis between residuals.

@@

In this case, the employment of \textit{auto.arima} stepwise methodology does not lead to any useful model, since it obtains $\mathrm{AICc}=\inf$ for the two manually-chosen models, that have an AICc value around -84.500, while the best model found by that methodology has an AICc around -81.000. This can be caused by approximation errors\footnote{The \textit{auto.arima} methodology was used only with initial approximation, due to high computing time}, as we supposed earlier.

@@

Analytically, we have quite a difference between the two models in terms of adherence of the confidence intervals:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item $\mathrm{ARIMA}(0,1,1)(0,1,1)_{168}$: $\approx98\%$ of data is in the $80\%$ CI and $\approx99\%$ in the $95\%$ CI
	\item $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$: $\approx73\%$ of data is in the $80\%$ CI and $\approx89\%$ in the $95\%$ CI
\end{itemize}
As we see, the first model over-generalises the variability, while the latter seems more precise, even if losing some information. Also considering the MAPE on the test set, the latter model fits the data better with a $16\%$ error, instead of a $18\%$.

Thus, we can select the $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$ model on the Box-Cox($\lambda\approx-1$)-transformed data, still being aware of the missed 24 hours seasonality, similarly to the Uccle station.

@@

\section{Conclusion}
\subsection{Lockdown influence}
Even if the two models have the same order, they cannot be directly compared, for the Membach model works after a transformation pass.

@@

\subsection{Limitations and future works}
The main limitations of this projects that we can identify are: multiple seasonalities and Normality assumption.

>>>>

\subsubsection{Motivation and objective}
Thanks to the recent COVID-19 outbreak worldwide, we currently have data about many different phoenomena from periods with standard human activity, and from a period with reduced human activity.

@@

It is important to notice that the daily data in each file refers to the UTC time instead of the local Europe/Brussels time, an optimal choice to publish data in a machine-readable way, even if this could cause some issues while interpreting the results.

@@

\subsubsection{Data aggregation and initial transformation}
Since we are not interested in a second-precise analysis, and it would be computationally expensive to deal with the whole dataset, we are going to aggregate data by hour. The aggregation will get us around 14 thousand points, that can still result in interpretable models.

@@

In particular, after a visual analysis of the decomposition results, we can suppose the 90 hours gap from the Uccle station is related to some changes to the seismic equipment, that resulted in slightly more precise measurements.

@@

Since our main focus is on the seasonalities themselves, we aim to prevent assumptions based on this objective. In other words, we want to avoid searching for correlation with human activity while imposing human-based seasonality periods. Thus, we will use an analytical method to extract seasonalities that the data itself suggests to us. Still, we recognize that the already performed transformations impose an human-based time alteration to the data itself (Section \ref{sec:cleanup-transform}).

@@

On this note we must specify that, when a significant periodicity is found, it influences smaller periodicities given by $f*\{2,3,\dots\}$ the significant frequency. Also, if any periodicity is found that is longer than the original dataset, it must be considered as simple chance, rather than serendipity.

@@

In \figurename~\ref{fig:decomposed} we see the decomposition results. While data from Uccle station have a clear significativity of both trend and seasonalities, data from Membach could seem not significant. Still, if we notice that the main timeseries lies between values of 0 and 100, we can find out that around half of the value will be composed by the trend, a 30\% by the joined seasonalities component, and the remaining value by the residuals. We will need to better handle the spikes starting from the middle of our dataset.

@@

Some R commands cannot handle timeseries with missing values, but as we have seen both stations datasets include missing values.
Also, please note that we are going to have a timeseries object indexed by weeks, so that it contains the 168 hours between each integer.
Since we know that the longest missing values series is 90 hours long, we can apply a simple completion strategy for serving those commands a filled timeseries: we will substitute missing values with the mean of the week they are in; in R terms \texttt{na.aggregate(x, floor)}.

@@

Since our dataset shows a relevant trend component, resulting in non-constant mean, we consider the usage of the \textit{diff}\footnote{The \textit{diff} operator applies the \textit{d}-order differences, by default on lag 1 (\textit{simple}). All \textit{diff} operators we will apply use $d=1$.} operator to smooth the mean.
After applying the \textit{diff} operator with lag 168, in conjunction with the simple \textit{diff} or not, we come up with two possible models by checking the ACF and PACF plots:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item from using both \textit{diff}s, the ACF and PACF on the long run hint to $\mathrm{ARIMA}(0,1,0)(0,1,1)_{168}$
	\item from using just the lagged \textit{diff}, the PACF on the first lags and the ACF/PACF combination on the long run hint to $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$
\end{itemize}
We also employ the \textit{auto.arima} stepwise methodology\footnote{The \textit{auto.arima} stepwise methodology fits different ARIMA models within a range of orders, and based on each result's information criteria it will try to minimize it by changing each time one of the \textit{p}, \textit{d}, \textit{q}, \textit{P}, \textit{D}, or \textit{Q} values, depending on the configured limits.}, locking the \textit{diff} values to 0 or 1, to get some hints on the possible models.

@@

\begin{figure}[h]
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-26} %30
		\caption{ACF and PACF plots}
		\label{fig:arima-26}
	\end{subfigure}
	\begin{subfigure}{.5\linewidth}
		\includegraphics[width=\linewidth]{project.uccs_files/figure-latex/arima-24} %28
		\caption{Q-Q plot}
		\label{fig:arima-24}
	\end{subfigure}
	\begin{center}
		\caption{Plots for $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$}
		\label{fig:arima}
	\end{center}
\end{figure}

The ACF and PACF of the residuals for all three models are really similar (\figurename~\ref{fig:arima-26}), even if the models themselves have great differences in meaning and in number of parameters: they highlight a clear seasonal ARMA process with periodicity 24. This shows how our previous conclusion that the 168 hours seasonality could include the 24 hours one is flawed. Unfortunately, the parameters estimator we are using does not permit to estimate multiple seasonalities\footnote{\textit{Arima} method from the \textit{forecast} R package, only specifying the seasonal and non-seasonal orders.}.

Probably related to the 24 hours periodicity that remains on the residuals, together with the non-Normality of the original data, all Q-Q plots show clear divergence on both sides of the distribution (\figurename~\ref{fig:arima-24}). The Box-Pierce test statistics accept the independence hypothesis between residuals.

@@

In this case, the employment of \textit{auto.arima} stepwise methodology does not lead to any useful model, since it obtains $\mathrm{AICc}=\inf$ for the two manually-chosen models, that have an AICc value around -84.500, while the best model found by that methodology has an AICc around -81.000. This can be caused by approximation errors\footnote{The \textit{auto.arima} methodology was only used with initial approximation, due to high computing time.}, as we supposed earlier.

@@

Analytically, we have quite a difference between the two models in terms of adherence of the confidence intervals:
\begin{itemize}[topsep=0.5em,itemsep=0em,partopsep=0.5em]
	\item $\mathrm{ARIMA}(0,1,1)(0,1,1)_{168}$: $\approx98\%$ of data is in the $80\%$ CI and $\approx99\%$ in the $95\%$ CI
	\item $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$: $\approx73\%$ of data is in the $80\%$ CI and $\approx89\%$ in the $95\%$ CI
\end{itemize}
As we see, the first model over-generalises the variability, while the latter seems more precise, even if losing some information. Also considering the MAPE on the test set, the latter model fits the data better with a $16\%$ error, instead of the $18\%$ for the first model.

Thus, we can select the $\mathrm{ARIMA}(2,0,0)(0,1,1)_{168}$ model on the Box-Cox($\lambda\approx-1$) transformed data, still being aware of the missed 24 hours seasonality, similarly to the Uccle station.

@@

\section{Conclusion}
\subsection{Lockdown influence}
Even if the two models have the same order, they cannot be directly compared using the Information Criteria, for the Membach model works after a transformation pass, and they are fitted from different data.

@@

\subsection{Limitations and future works}
The main problems of this projects that we identified are multiple seasonalities and the Normality assumption.
